SILVR: Conditional diffusion model for molecule generation without additional training

### 2023-03-16 paper/manuscript.tex
pushing progress on theory

%%%%
% Subsection: SILVR
%%%%
\subsection{Selective Iterative Latent Variable Refinement}
We propose a new method, that combines the iterative variable refinement (ILVR) by Choi et al.~\cite{choi2021ilvr} and take it from the image generation process and incorporate it into the EDM by Hoogeboom et al. to produce a way to generate a selective iterative laten variable refinement (SILVR) procedure in which we use a reference molecule to introduce into the denoising step. The reference used can be a small fragment, multiple fragments or even include dummy atoms that do not exist in the original reference. As a result we propose a modification of Algorithm 2 by Hoogeboom et al.~\cite{hoogeboom2022equivariant} to result in the SILVR algorithm~\ref{alg:silvr}.  We are considering latent space variables $\mathbf{z}=[\mathbf{z}^{(x)},\mathbf{z}^{(h)}]$ for the standard denoising process and  $\mathbf{\hat{z}} = [\mathbf{\hat{z}}^{(y)},\mathbf{\hat{z}}^{(h)}]$, for the set of reference coordinates given by $\mathbf{y}$. The vector $\mathbf{h}$ contains all the scalar node properties of the equivariant EDM. 
\begin{algorithm}
\caption{SILVR}\label{alg:silvr}
\begin{algorithmic}[1]
\State \textbf{Input:} Reference molecule $\mathbf{y,h}$, neural network $\phi$ 
\State \textbf{Output:} Generated molecule $\mathbf{x,h}$

\State Subtract centre of gravity (cg) from $\mathbf{y,h}$
\State Sample $z_T \sim \mathcal{N}(\mathbf{0,I})$
\For {$t$ = $T,...,1$}
    \State Sample $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0,I})$
    \State Subtract centre of gravity from $\boldsymbol{\epsilon}^{(x)}$
    \State Get reference to right noise level $\mathbf{\hat{z}}_t = \alpha_t \mathbf{\hat{z}}_{t=0} +\sigma_t\times\boldsymbol{\epsilon}$
    \State Compute $\mathbf{z}_{t-1}$ using $\phi(\mathbf{z}_t,t)$
    \State Update $\mathbf{z}_{t-1} \gets \mathbf{z}_{t-1} - \alpha_{t-1}r_{\mathrm{S}} \mathbf{z}_{t-1} +r_{\mathrm{S}} \mathbf{\hat{z}}_{t-1}$ \Comment{SILVR equation}
    \State Subtract centre of gravity from updated $\mathbf{z}_{t-1}$

\EndFor
\State Sample $\mathbf{x,h} \sim p(\mathbf{x,h|z_0})$ 
\end{algorithmic}
\end{algorithm}

@@

The core of the new method is the addition of a refinement process within the denoising part during run time. The resulting SILVR model produces conditional samples without any conditional training when generating new molecules. To illustrate the usefulness of this run-time modification, we show how SILVR can be used in the context of fragment-based drug design. The goal is to produce molecules that resemble a binding site based on exiting fragments and their set of atomic coordinates.

\section{Methods}
As stated before, the selective iterative latent variable refinement (SILVR) method makes use of the pre-trained equivariant diffusion model (EDM) as introduced by Hoogeboom \textit{et al.}~\cite{hoogeboom2022equivariant}.

\subsection{Conditioning the pre-trained EDM with SILVR}
The EDM by Hoodgeboom \textit{et al.} was trained on the 30 lowest energy conformations of 430,000 molecules from the GEOM dataset with explicit protons. A simplified schematic describing the EDM architecture is shared in scheme~\ref{fig:_architecture} for more details on the model theory and training refer to Hoogeboom et al.~\cite{hoogeboom2022equivariant}. This model strictly only considers atomic coordinates; all bond information is ignored. A more recent version of the EDM also accounts for bond order and SILVR could also be incorporated into this new model.~\cite{hvignac2023midi}. During training, the atomic coordinates (and their element) are passed through a forward diffusion process with iterative addition of Gaussian noise, where the extent of noise added at each step is defined by parameter $\beta$ (N.B. scheme X shows the diffusion process as a Markov chain, however in practice the state at time $t$ is determined as a direct transformation of the initial state). The diffusion process is eventually terminated when $t=T$, by which point all structure is lost and all coordinates follow a Gaussian distribution. An equivariant graph neural network (EGNN) is then trained to predict the reverse process, predicting the previous state in the sequence given any state. At run time, the generative model is instantiated with a sample from a Gaussian distribution and the series of denoising steps are applied resulting in a generated sample consisting of atomic coordinates resembling a drug-like molecule. The XYZ coordinates can then be interpreted using cheminformatics software to determine the molecular graph.

The EDM was then adapted by introducing SILVR within the denoising process algorithm~\ref{alg:silvr}. At run time, each atom of the reference set of coordinates is mapped to an atom in the EDM. The reference coordinates are then translated such that their center of mass is at the origin. The reference coordinates are then diffused to the same timestep as that of the denoising process. That is, the amount of structure remaining from the reference should match the amount of structure formed by the generative process. A small refinement is applied to add information from the reference coordinates to the latent variable of the denoising process~\ref{eq:denoising}. This equation has the effect of shrinking the coordinates towards the origin, and then expanding the coordinates out in the direction of the reference. Importantly, the extent of this refinement is defined by the SILVR rate $r_{\mathrm{S}}$, with $r_{\mathrm{S}}=0$ providing no additional refinement and $r_{\mathrm{S}}=1$ will result in a total replacement of atoms. The diffusion of the reference to $t=\tau$ requires the sampling of a Gaussian; while in principle this sampling could be performed only once prior to the denoising loop, in practice better results were obtained when the Gaussian was repeatedly sampled during the denoising process. Once denoising is complete, sampled molecules can be translated back to the same coordinates as the reference by reverting the initial centre of gravity transformation; in the case of fragment data, this has the effect of returning samples to the binding site of the protein.

@@

\subsection{Reference Dataset: Covid-moonshot}
Reference molecules were selected from the original 23 non-covalent hits of the SARS-CoV-2 Main Protease (Mpro) identified by the XChem fragment screen as part of the Covid Moonshot Project~\cite{moonshot}. A more detailed picture of all fragements is presented in Figure~\ref{si:figure}. Fragments were visualised using XXX v XXX and combinations were arbitrarily selected as test cases. Fragments \texttt{x0072} and \texttt{x0354} were selected for benchmarking the effect of $r_{\mathrm{S}}$ on sampling; \texttt{x1093}, \texttt{x0072}, \texttt{x2193} were selected to represent three significantly overlapping fragments; \texttt{x0434}, \texttt{x0305} and \texttt{x0072}, \textit{x0354} were selected as partially overlapping fragments; and \texttt{x0874}, \texttt{x0397} were selected as two disconnected fragments, resembling a linker design type problem. The bonding information of selected fragments was deleted and XYZ coordinates were combined into a single XYZ file. Values of $r_{\mathrm{S}}$ were selected and added to the XYZ file to create a reference file containing all experiment setup information. Each experiment was sampled 1000 times.

\subsection{Diferent observables were monitored to gauge the success of computational experiments for the generation of new models}

Different observables were used to monitor how realistic and reliable new molecules were generated and how well they can fit into the existing binding site of the SARS-CoV-2 main protease.

\subsubsection{Atom stability}
The accuracy of placement of atoms was determined using the stability metric proposed by Hoogeboom \textit{et al.}. This metric infers bonds and bond orders between atoms by considering their interatomic distances. Once all bonds are defined, the valance of each atom is compared to its expected valance and if these values match, the atom is determined as stable. It should be noted that this metric requires the presence of all hydrogens (both explicit and implicit) for an atom to be decided as stable. For comparability with other similar published models, this metric was used unmodified. The additional metric of “molecular stability” is often reported together with atom stability (if every atom is stable then the whole molecule is stable), however as has been previously identified, large molecules sampled from this GEOM-trained EDM tend to be unstable.

\subsubsection{SILVR to reference}
The SILVR algorithm creates a one-to-one mapping between reference atom coordinates and sample atom coordinates. The RMSD for this pairwise mapping, ignoring atom identities, was calculated to determine the spacial similarity of samples to the reference.

\subsubsection{Geometric stability - Auto3D}
To determine whether samples represent a true molecular geometry, independent minimisation of molecular geometries was performed using Atoms In Molecules Neural Network Potential (AIMNet) with Auto3D. All samples were read by RDKit and samples containing more than one molecule were removed from the test set. The SMILES string of each molecule was written to a new file, read by Auto3D, and geometry predicted by AIMNet. The RMSD between SILVR-generated coordinates and the Auto3D minimised coordinates were calculated with RDKit.

@@

\subsubsection{Maybe: Docking of molecules}

>>>>

%%%%
% Subsection: SILVR
%%%%
\subsection{Selective Iterative Latent Variable Refinement}
\label{sec:silvr}
We propose a new method, that combines the iterative variable refinement (ILVR) by Choi et al.~\cite{choi2021ilvr} proposed in an image generation context and incorporates the ideas into the EDM by Hoogeboom et al.. This allows us to generate a selective iterative laten variable refinement (SILVR) procedure in which we introduce a reference molecule into the denoising step. The reference used can be a small fragment, multiple fragments or even include dummy atoms that do not exist in the original reference. As a result we propose a modification of Algorithm 2 by Hoogeboom et al.~\cite{hoogeboom2022equivariant} to result in the SILVR algorithm~\ref{alg:silvr}.  We are considering latent space variables $\mathbf{z}=[\mathbf{z}^{(x)},\mathbf{z}^{(h)}]$ for the standard denoising process and  $\mathbf{\hat{z}} = [\mathbf{\hat{z}}^{(y)},\mathbf{\hat{z}}^{(h)}]$, for the set of reference coordinates given by $\mathbf{y}$. The vector $\mathbf{h}$ contains all the scalar node properties of the equivariant EDM. 
\begin{algorithm}
\caption{SILVR}\label{alg:silvr}
\begin{algorithmic}[1]
\State \textbf{Input:} Reference molecule $\mathbf{y,h}$, EDM $\kappa$ 
\State \textbf{Output:} Generated molecule $\mathbf{x,h}$

\State Compute $\mathbf{\hat{z}}_0$ from $\mathbf{y, h}$, such that $[\mathbf{z}^{(y)} , \mathbf{z}^{(h)} ] = f (\mathbf{y, h})$ is E(3). 
\State Subtract center of gravity (COG) from $\mathbf{\hat{z}}_0$
\State Sample $z_T \sim \mathcal{N}(\mathbf{0,I})$
\For {$t$ = $T,...,1$}
    \State Sample $\boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0,I})$
    \State Subtract COG from $\boldsymbol{\epsilon}^{(x)}$
    \State $\mathbf{\hat{z}}_{t-1} = \alpha_{t-1} \mathbf{\hat{z}}_{0} +\sigma_{t-1}\times\boldsymbol{\epsilon}$
    \Comment{Noise reference to t=t-1}
    \State $\mathbf{z}_{t-1} \gets \kappa (\mathbf{z}_t,t)$
    \Comment{Compute denosing step}
    \State Update $\mathbf{z}_{t-1} \gets \mathbf{z}_{t-1} - \alpha_{t-1}r_{\mathrm{S}} \mathbf{z}_{t-1} +r_{\mathrm{S}} \mathbf{\hat{z}}_{t-1}$ \Comment{SILVR equation}
    \State Subtract COG from updated $\mathbf{z}_{t-1}$

\EndFor
\State Add COG (from line 4) to $\mathbf{z}_0$
\Comment{Move sample back to pocket}
\State Sample $\mathbf{x,h} \sim p(\mathbf{x,h|z_0})$ 
\end{algorithmic}
\end{algorithm}

@@

The core of the new method is the addition of a refinement process within the denoising part during run time. The resulting SILVR model produces conditional samples without any conditional training when generating new molecules. Figure~\ref{fig:silvr_explanation} shows an illustrative example of how the SILVR rate $r_{\mathrm{S}}$ is used to shift the latent space vector $\mathbf{z}_{t-1}$ at any point in the denoising process from $T\ldots t=1$. Using the silver equation as defined in the SILVR algorithm~\ref{alg:silvr} line 11, an existing denoising step (green) is brought closer to the reference (purple) in latent space.

\section{Methods}
To illustrate the usefulness of this run-time modification, we show how SILVR can be used in the context of fragment-based drug design. The goal is to produce molecules that resemble a binding site based on exiting fragments and their set of atomic coordinates. 
%As stated before, the selective iterative latent variable refinement (SILVR) method makes use of the pre-trained equivariant diffusion model (EDM) as introduced by Hoogeboom \textit{et al.}~\cite{hoogeboom2022equivariant}.

\subsection{Conditioning the pre-trained EDM with SILVR}
The EDM by Hoodgeboom \textit{et al.} was trained on the 30 lowest energy conformations of 430,000 molecules from the GEOM dataset with explicit protons. A simplified schematic describing the EDM architecture is shared in scheme~\ref{fig:_architecture} for more details on the model theory and training refer to Hoogeboom et al.~\cite{hoogeboom2022equivariant}. This model strictly only considers atomic coordinates; all bond information is ignored. A more recent version of the EDM also accounts for bond order and SILVR could also be incorporated into this new model~\cite{hvignac2023midi}. During training, the atomic coordinates (and their element) are passed through a forward diffusion process with iterative addition of Gaussian noise, where the extent of noise added at each step is defined by parameter $\beta$ (N.B. scheme~\ref{fig:schematic} shows the diffusion process as a Markov chain, however in practice the state at time $t$ is determined as a direct transformation of the initial state). The diffusion process is eventually terminated when $t=T$, by which point all structure is lost and all coordinates follow a Gaussian distribution. An equivariant graph neural network (EGNN) is then trained to predict the reverse process, predicting the previous state in the sequence given any state. At run time, the generative model is instantiated with a sample from a Gaussian distribution and the series of denoising steps are applied resulting in a generated sample consisting of atomic coordinates resembling a drug-like molecule. The \texttt{XYZ} coordinates can then be interpreted using cheminformatics software to determine the molecular graph.

The EDM was then adapted by introducing SILVR within the denoising process algorithm~\ref{alg:silvr}, as outlined in~\ref{sec:silvr}. At run time, each atom of the reference set of coordinates is mapped to an atom in the EDM. The reference coordinates are then translated such that their center of mass is at the origin. The reference coordinates are then diffused to the same timestep as that of the denoising process. That is, the amount of structure remaining from the reference should match the amount of structure formed by the generative process. A small refinement is applied to add information from the reference coordinates to the latent variable of the denoising process~\ref{eq:denoising}. This equation has the effect of shrinking the coordinates towards the origin, and then expanding the coordinates out in the direction of the reference, see Figure~\ref{fig:silvr_explanation}. Importantly, the extent of this refinement is defined by the SILVR rate $r_{\mathrm{S}}$, with $r_{\mathrm{S}}=0$ providing no additional refinement and $r_{\mathrm{S}}=1$ will result in a total replacement of atoms. The diffusion of the reference to $t=\tau$ requires the sampling of a Gaussian; while in principle this sampling could be performed only once prior to the denoising loop, in practice better results were obtained when the Gaussian was repeatedly sampled during the denoising process. Once denoising is complete, sampled molecules can be translated back to the same coordinates as the reference by reverting the initial centre of gravity transformation; in the case of fragment data, this has the effect of returning samples to the binding site of the protein.

@@

\subsection{Reference Dataset: Covid-moonshot}
Reference molecules were selected from the original 23 non-covalent hits of the SARS-CoV-2 Main Protease (Mpro) identified by the XChem fragment screen as part of the Covid Moonshot Project~\cite{consortium2023open, consortium2021open}. A more detailed picture of all fragements is presented in Figure S1 of the SI~\ref{si}. Fragments were visualised using NGLview version 3.03 and combinations were arbitrarily selected as test cases. Fragments \texttt{x0072} and \texttt{x0354} were selected for benchmarking the effect of $r_{\mathrm{S}}$ on sampling; \texttt{x1093}, \texttt{x0072}, \texttt{x2193} were selected to represent three significantly overlapping fragments; \texttt{x0434}, \texttt{x0305} and \texttt{x0072}, \texttt{x0354} were selected as partially overlapping fragments; and \texttt{x0874}, \texttt{x0397} were selected as two disconnected fragments, resembling a linker design type problem. The bonding information of selected fragments was deleted and \texttt{XYZ} coordinates were combined into a single \texttt{XYZ} file. Values of $r_{\mathrm{S}}$ were selected and added to the \texttt{XYZ} file to create a reference file containing all experiment setup information. Each experiment was sampled 1000 times.

\subsection{Different observables were monitored to gauge the success of computational experiments for the generation of new models}

Different observables were used to monitor how realistic and reliable newly generated molecules were and how well they can fit into the existing binding site of MPro. We looked at the following set of measures:

\subsubsection{Atom stability}
The accuracy of placement of atoms was determined using the stability metric proposed by Hoogeboom et al.. This metric infers bonds and bond orders between atoms by considering their interatomic distances. Once all bonds are defined, the valance of each atom is compared to its expected valance and if these values match, the atom is determined as stable. It should be noted that this metric requires the presence of all hydrogens (both explicit and implicit) for an atom to be decided as stable. For comparability with other similar published models, this measure was used unmodified. The additional measure of \textit{molecular stability} is often reported together with atom stability (if every atom is stable then the whole molecule is stable), however as has been previously identified, large molecules sampled from this GEOM-trained EDM tend to be unstable.

\subsubsection{SILVR to reference}
The SILVR algorithm creates a one-to-one mapping between reference atom coordinates and sample atom coordinates. The RMSD for this pairwise mapping, ignoring atom identities, was calculated to determine the spacial similarity of samples to the reference. All RMSD calculations were carried out using \url{https://github.com/charnley/rmsd} version 1.5.1.

\subsubsection{Geometric stability - Auto3D}
To determine whether samples represent a true molecular geometry, an independent minimisation of molecular geometries was performed using Atoms In Molecules Neural Network Potential (AIMNet) with Auto3D. All samples were read by RDKit and samples containing more than one molecule were removed from the test set. The SMILES string of each molecule was written to a new file, read by Auto3D, and geometry predicted by AIMNet. The RMSD between SILVR-generated coordinates and the Auto3D minimised coordinates were calculated with RDKit version 2022.03.5.

@@
